Reasoning About a Machine with Local Capabilities

### 2018-05-07 toplas/paper.tex
Add an intro to section 7 + some other minor changes.

\begin{toplas}
\section{Reasoning about programs on a capability machine}
\label{sec:reasoning}
% \itoplassug{
%   \begin{itemize}
%   \item A section with the lemmas we use for reasoning about programs
%     \begin{itemize}
%     \item lemma 58, scall works. Include description of how it handles the
%       common things in call and leaves the argumentation for callee code and
%       continuation.
%     \item lemma 59, malloc works. If this is included, we should probably also
%       include the malloc specification.
%     \item lemma 60, create closure works.
%     \item Include a description about how the FTLR is used to reason about
%       unknown code.
%     \item Mention that this is a step on the way towards having a program logic,
%       and then in the discussion discuss why a program logic for this language
%       is not what we need.
%     \end{itemize}
%   \end{itemize}
% }
\itoplassug{add actual intro to this section.}

@@

% Many details to get right when reasoning about low-level programs.
There are many details to get right when programming in assembly. These details
carry over to proofs about assembly programs, so the proofs become cumbersome.
It is especially annoying, when the same line of reasoning is applied in
multiple places. For instance, in $\texttt{scall}$ the setup before the jump to
the call is always the same and the recovery upon return is also always the
same, and in both cases the reasoning will be the same every time.
We deal with this part of the reasoning in the following lemma.
\begin{lemma}[$\mathtt{scall}$ works]
  \label{lem:scall-works}
  If
  \begin{enumerate}
  \item \label{item:scall:memsat} $\memSat[n]{\ms}{\revokeTemp{W}}$ 
  \item $\dom(\ms_f) \cap (\dom(\ms_\stk \uplus \ms_{\mathit{unused}} \uplus \ms)) = \emptyset$
  \item \label{item:scall:stack} $(\reg,\ms) \text{ is looking at }
    \mathtt{scall}\;r(\overline{r_{\mathit{arg}}},
    \overline{r_{\mathit{priv}}}) \text{ followed by }
    c_{\mathit{next}}$\footnote{Defined in the appendix.}
  \item $\reg \text{ points to stack with $\ms_\stk$ used and $\ms_{\mathit{unused}}$ unused}$
  \item \label{item:scall:hyp-callee} \emph{Hyp-Callee} If
    \begin{itemize}
    \item $\dom(\ms_{\mathit{unused}}) = \dom(\ms_{\mathit{rec}}
      \uplus \ms_{\mathit{unused}}')$,
    \item $W' =
      \revokeTemp{W}[\iota^{\sta}(\temp,\ms_\stk\uplus\ms_{\mathit{rec}} \uplus \ms_f),\iota^{\pwl}(\dom(\ms_{\mathit{unused}}'))]$,
    \item $\memSat[n-1]{\ms''}{W'}$
    \item $\reg' \text{ points to stack with $\emptyset$ used and $\ms_{\mathit{unused}}'$ unused}$
    \item $\reg'= \reg_0[\pcreg\mapsto\updatePcPerm{\reg(r)},
      \overline{r_{\mathit{arg}}} \mapsto \reg(\overline{r_{\mathit{arg}}}),r_0
      \mapsto c_{\mathit{ret}}, r_\stk \mapsto c_\stk, r \mapsto \reg(r)]$ 
    \item $\npair[n-1]{c_{\mathit{ret}}} \in \stdvr(W')$
    \item $\npair[n-1]{c_\stk} \in \stdvr(W')$
    \end{itemize}

@@

The \texttt{malloc} instruction is also used often, so we also prove
Lemma~\ref{lem:malloc-works} to help reason about it. The structure of the
\texttt{malloc} lemma is close to that of the \texttt{scall} lemma. It also has
requirements on the configuration just before \texttt{malloc} is executed
(\ref{item:malloc:looking-at}-\ref{item:malloc:malloc-sat}) as well as
requirements on the execution afterwards (\ref{item:malloc:hyp-cont}). It does
not have any requirements on the callee as the \texttt{malloc} specification
defines its behavior.
\begin{lemma}[$\mathtt{malloc}$ works]
  \label{lem:malloc-works}
  If
  \begin{enumerate}
  \item \label{item:malloc:looking-at}\lookingat{(\reg,\ms)}{\mathtt{malloc}\;r\;k}{c_{\var{next}}}
  \item $k \geq 0$
  \item \linksto{(\reg,\ms)}{\malloc}{k}{c_\malloc}
  \item $c_\malloc$ satisfies the $\malloc$ specification with $\iota_{\malloc,0}$
  \item $W \futurestr [i \mapsto\iota_{\malloc,0}]$
  \item $\memSat{\ms}{W}$
  \item $\ms = \ms' \uplus \ms_{\var{footprint}}$
  \item \label{item:malloc:malloc-sat} $\memSat{\ms_{\var{footprint}}}{[i \mapsto W(i)]}$
  \item \label{item:malloc:hyp-cont} \emph{Hyp-Cont}
    If
    \begin{itemize}
    \item $n' \leq n-1$
    \item $\iota_\malloc \futurewk W(i)$
    \item $\memSat[n']{\ms_{\var{footprint}}' \uplus \ms'}{W[i \mapsto \iota_\malloc]}$
    \item $\memSat[n']{\ms_{\var{footprint}}'}{[i \mapsto \iota_\malloc]}$
      \begin{equation*}
        \reg'(r') = 
        \begin{cases}
          c_{\var{next}} & r' = \pcreg \\
          ((\rwx,\glob),\start,\addrend,\addr) & r' = r \\
          \reg(r) & r' \not\in \RegName_t \union \{\pcreg, r, r_1\}
        \end{cases}
      \end{equation*}
    \item $\addrend - \start = k - 1$
    \item $\dom(\ms_{\var{alloc}}) = [\start,\addrend]$
    \item $\forall \addr \in [\start,\addrend]\ldotp \ms_{\var{alloc}}(\addr) = 0$
    \end{itemize}
    Then we have $\npair[n']{(\reg',\ms' \uplus \ms_{\var{footprint}}'  \uplus \ms_{\var{alloc}})} \in \observations(W[\iota_\malloc])$
  \end{enumerate}
  Then
  \[
    \npair{(\reg,\ms)} \in \observations(W)
  \]
\end{lemma}
In the technical appendix~\citep{technical_appendix}, we also define a lemma for
the macro used to create closures, \texttt{crtcls}.
% The below remark seems unnecessary: isn't this true about any lemma in any proof?
% We stress that these lemmas
% are conveniences and are not strictly necessary to reason about programs for the
% capability machine. In principle, one could reason solely in the semantic model
% without the use of these lemmas, but that would lead to a lot of repetition.
\end{toplas}

>>>>

\begin{toplas}
\section{Reasoning about programs on a capability machine}
\label{sec:reasoning}
% \itoplassug{
%   \begin{itemize}
%   \item A section with the lemmas we use for reasoning about programs
%     \begin{itemize}
%     \item lemma 58, scall works. Include description of how it handles the
%       common things in call and leaves the argumentation for callee code and
%       continuation.
%     \item lemma 59, malloc works. If this is included, we should probably also
%       include the malloc specification.
%     \item lemma 60, create closure works.
%     \item Include a description about how the FTLR is used to reason about
%       unknown code.
%     \item Mention that this is a step on the way towards having a program logic,
%       and then in the discussion discuss why a program logic for this language
%       is not what we need.
%     \end{itemize}
%   \end{itemize}
% }
There are many details to get right when programming in assembly. These details
carry over to proofs about assembly programs, so many of the proofs in the next
section about example programs are a bit cumbersome. It is especially annoying,
when the same line of reasoning is applied in multiple places. To mitigate this,
we have defined a number of lemmas that capture common reasoning patterns in
these proofs. We present the most important such lemmas in this section, to give
the reader an idea of how they work.

@@

% Many details to get right when reasoning about low-level programs.
Another common pattern in proofs about programs on our capability machine
concerns the use of scall. The following lemma captures the commonalities of
reasoning about programs using scall. From another point of view, it can be seen
as a specification for scall.
\begin{lemma}[$\mathtt{scall}$ works]
  \label{lem:scall-works}
  If
  \begin{enumerate}
  \item \label{item:scall:memsat} $\memSat[n]{\ms}{\revokeTemp{W}}$ 
  \item $\dom(\ms_f) \cap (\dom(\ms_\stk \uplus \ms_{\mathit{unused}} \uplus \ms)) = \emptyset$
  \item \label{item:scall:stack} $(\reg,\ms) \text{ is looking at }
    \mathtt{scall}\;r(\overline{r_{\mathit{arg}}},
    \overline{r_{\mathit{priv}}}) \text{ followed by }
    c_{\mathit{next}}$\footnote{Defined in the appendix.}
  \item $\reg \text{ points to stack with $\ms_\stk$ used and $\ms_{\mathit{unused}}$ unused}$
  \item \label{item:scall:hyp-callee} \emph{Hyp-Callee} If
    \begin{itemize}
    \item $\dom(\ms_{\mathit{unused}}) = \dom(\ms_{\mathit{rec}}
      \uplus \ms_{\mathit{unused}}')$,
    \item $W' =
      \revokeTemp{W}[\iota^{\sta}(\temp,\ms_\stk\uplus\ms_{\mathit{rec}} \uplus \ms_f),\iota^{\pwl}(\dom(\ms_{\mathit{unused}}'))]$,
    \item $\memSat[n-1]{\ms''}{W'}$
    \item $\reg' \text{ points to stack with $\emptyset$ used and $\ms_{\mathit{unused}}'$ unused}$
    \item $\reg'= \reg_0[\pcreg\mapsto\updatePcPerm{\reg(r)},
      \overline{r_{\mathit{arg}}} \mapsto \reg(\overline{r_{\mathit{arg}}}),r_0
      \mapsto c_{\mathit{ret}}, r_\stk \mapsto c_\stk, r \mapsto \reg(r)]$ 
    \item $\npair[n-1]{c_{\mathit{ret}}} \in \stdvr(W')$
    \item $\npair[n-1]{c_\stk} \in \stdvr(W')$
    \end{itemize}

@@

Also often used is the \texttt{malloc} instruction, so we also prove
Lemma~\ref{lem:malloc-works} to help reason about it. The structure of the
\texttt{malloc} lemma is close to that of the \texttt{scall} lemma. It also has
requirements on the configuration just before \texttt{malloc} is executed
(\ref{item:malloc:looking-at}-\ref{item:malloc:malloc-sat}) as well as
requirements on the execution afterwards (\ref{item:malloc:hyp-cont}). It does
not have any requirements on the callee as the \texttt{malloc} specification
defines its behavior.
\begin{lemma}[$\mathtt{malloc}$ works]
  \label{lem:malloc-works}
  If
  \begin{enumerate}
  \item \label{item:malloc:looking-at}\lookingat{(\reg,\ms)}{\mathtt{malloc}\;r\;k}{c_{\var{next}}}
  \item $k \geq 0$
  \item \linksto{(\reg,\ms)}{\malloc}{k}{c_\malloc}
  \item $c_\malloc$ satisfies the $\malloc$ specification with $\iota_{\malloc,0}$
  \item $W \futurestr [i \mapsto\iota_{\malloc,0}]$
  \item $\memSat{\ms}{W}$
  \item $\ms = \ms' \uplus \ms_{\var{footprint}}$
  \item \label{item:malloc:malloc-sat} $\memSat{\ms_{\var{footprint}}}{[i \mapsto W(i)]}$
  \item \label{item:malloc:hyp-cont} \emph{Hyp-Cont}
    If
    \begin{itemize}
    \item $n' \leq n-1$
    \item $\iota_\malloc \futurewk W(i)$
    \item $\memSat[n']{\ms_{\var{footprint}}' \uplus \ms'}{W[i \mapsto \iota_\malloc]}$
    \item $\memSat[n']{\ms_{\var{footprint}}'}{[i \mapsto \iota_\malloc]}$
      \begin{equation*}
        \reg'(r') = 
        \begin{cases}
          c_{\var{next}} & r' = \pcreg \\
          ((\rwx,\glob),\start,\addrend,\addr) & r' = r \\
          \reg(r) & r' \not\in \RegName_t \union \{\pcreg, r, r_1\}
        \end{cases}
      \end{equation*}
    \item $\addrend - \start = k - 1$
    \item $\dom(\ms_{\var{alloc}}) = [\start,\addrend]$
    \item $\forall \addr \in [\start,\addrend]\ldotp \ms_{\var{alloc}}(\addr) = 0$
    \end{itemize}
    Then we have $\npair[n']{(\reg',\ms' \uplus \ms_{\var{footprint}}'  \uplus \ms_{\var{alloc}})} \in \observations(W[\iota_\malloc])$
  \end{enumerate}
  Then
  \[
    \npair{(\reg,\ms)} \in \observations(W)
  \]
\end{lemma}
In the technical appendix~\citep{technical_appendix}, we also define a lemma for
the macro used to create closures, \texttt{crtcls}.
% The below remark seems unnecessary: isn't this true about any lemma in any proof?
% We stress that these lemmas
% are conveniences and are not strictly necessary to reason about programs for the
% capability machine. In principle, one could reason solely in the semantic model
% without the use of these lemmas, but that would lead to a lot of repetition.
\end{toplas}
